Case Study: Gene Set Enrichment Analysis (Mouse Data Set)

### 2018-03-26 datasets/GSEA/gsea-human.Rmd

**Some sort of summary of the analysis.**

@@

**Some sort of summary of the data set.**  
We download a *DESeqDataSet* if not present locally.

```{r, message=FALSE}
if (!file.exists(count_file)) {
    download.file("https://www.dropbox.com/s/itrs2a6cnmurnck/human-counts.rds?dl=1",
                  destfile = count_file)
}
dsdObject <- readRDS(count_file)
```

We only perform differential tesing on protein coding genes, as specified in Ensembl release 86.

@@

**Add description of what's going on.**
We first test for differentially expressed genes using DESeq2 and use genes with adjusted p-value less than 0.1.

@@

We next determine the gene sets associated with each gene.

@@

Now we are ready to perform gene set enrichment analysis using `goseq`.

@@

**Add description of independent covariate.**

@@

@@

Generating the *SummarizedBenchmark* object:

@@

# Data Analysis (Null Comparison)

**Add description.**

## Create Null Data

```{r, eval=FALSE }
nms <- names( genes )
xx <- seq_len(length(nms))
gns <- 1*( xx %in% sample( xx, 1000, replace=FALSE ) )
names( gns ) <- nms
pwf <- nullp( gns, bias.data=geneLength )
goRes <- goseq( pwf, gene2cat=goSets )
res <- goRes %>%
    dplyr::filter( numDEInCat > 0 ) %>%
    dplyr::select( c("over_represented_pvalue", "numInCat") ) %>%
    dplyr::rename( pval = over_represented_pvalue, ind_covariate = numInCat )
```

```{r}
pr <- buildBench(bd, res, ftCols = "ind_covariate")
assayNames(pr) <- "qvalue"
pr <- addDefaultMetrics(pr)
```

>>>>

The objective of this document is to test different multiple testing methods in the context of Gene Set Enrichment Analysis (GSEA). In order to get a list of genes, we will use RNA-seq data consisting on cortex and cerebellum samples of a subset of the GTEx data.

@@

The dataset consists of 10 samples (6 cerebellum and 4 cortex samples), all from male individuals. We download a *DESeqDataSet* if not present locally and modify the design to test for differences in gene expression between cerebellums and cortex.

```{r, message=FALSE}
if (!file.exists(count_file)) {
    download.file("https://www.dropbox.com/s/itrs2a6cnmurnck/gsea-human-counts.rds?dl=1",
                  destfile = count_file)
}
dsdObject <- readRDS(count_file)
design( dsdObject ) <- ~ tissue
```

To keep running times short, we only perform differential tesing on protein coding genes, as specified in Ensembl release 86.

@@

Next, we run DESeq2 to retrieve a list of differentially expressed genes at a FDR of 10%.

@@

Next we’ll use the biomaRt package to download and associate GO annotations for each gene.

@@

Finally, we’ll perform a gene set enrichment analysis with goseq. goseq performs GSEA and adjusts for gene length biases. The bias is represented in the plot below, with gene length bin on the x-axis and proportion of genes detected to be differentially expressed on the y-axis.

@@

We will explore whether the size of the gene set can be used as a covariate for modern multiple-testing correction methods in the context of GSEA. In the plot below, the log10 of the p-values is plotted as a function of the size of the gene set. There is a pattern in which gene sets with a higher number of genes tend to have smaller p-values, which is indicative that gene set size is an informative covariate.

@@

We can also explore if the covariate seems to be independent under the null.

@@

We then execute the benchDesign and generate a *SummarizedBenchmark* object containing multiple-testing corrections using several methods.

@@

### 2018-03-26 datasets/GSEA/gsea-mouse.Rmd

**Some sort of summary of the analysis.**
Same analysis but using the HSC-MPP dataset from Nina.  
**Who is Nina?**

@@

Download *DESeqDataSet* if not present locally.

@@

**Why no filtering on gene_biotype == "protein_coding" for this data?**

@@

**Add description of what's going on.**
We first test for differentially expressed genes using DESeq2 and use genes with adjusted p-value less than 0.1.

@@

We next determine the gene sets associated with each gene.

@@

Here, we can see that the covariant variable is actually informative.

@@

**Add description of independent covariate.**

@@

>>>>

The objective of this vignette is to test different multiple testing methods in the context of Gene Set Enrichment Analysis (GSEA). To do this, we will use mouse data from the paper by [Cabezas-Wallscheid et al. (Cell stem Cell, 2014)](https://www.sciencedirect.com/science/article/pii/S1934590914003014?via%3Dihub). The dataset consists of RNA-seq data from mouse hematopoietic stem cells and 4 multipotent progenitor lineages. We will test for difference in gene expression between two different cell fractions and perform GSEA in the resulting list of genes.

@@

The data has been preprocessed and saved as a DESeqDataset object. The following lines of code download this *DESeqDataSet* if it is not present locally.

@@

**Why no filtering on gene_biotype == "protein_coding" for this data? AR: I think I did this in the other vignette to keep running times short. This shouldn't change the result since anyway protein coding genes are those with GO categories and genes without GO assignments are ignored.**

@@

In order to get a list of genes to test for GSEA, we will test each gene for differential gene expression between hematopoietic stem cells and multipotent progenitors (fraction 1). To do this, we will run DESeq2 to retrieve genes differentially expressed at a FDR of 10%.

@@

We next use biomaRt to get the relations between GO categories and genes.

@@

Here, we can see that the size of the gene set is actually informative.

@@

In the following plot, we explore the relationship between the p-value and the gene set size. We can see that this covariate is actually informative.

@@

And at the same time, it seems to be fair to assume that it is independent under the null hypothesis.
